
\documentclass[12pt,reqno]{report}      % Sets 12pt font, equation numbers on right
\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages
\usepackage{graphics}                 % Packages to allow inclusion of graphics
\usepackage{color}                    % For creating coloured text and background
\usepackage{siunitx}                  % for degrees celsius
\usepackage[toc,page]{appendix}
%\usepackage[margin=2in]{geometry}
\usepackage{lscape}
\usepackage{float}


\usepackage[colorlinks,citecolor=blue,linkcolor=blue]{hyperref}                 % For creating hyperlinks in cross references. It should be after the color package. The option colorlinks produces colored entries without boxes. The option citecolor=blue changes the default green citations to blue.
\renewcommand{\theequation}{\thechapter.\arabic{equation}}

\DeclareGraphicsExtensions{.pdf}
\parindent 0cm
\parskip 0.2cm
\topmargin 0.2cm
\oddsidemargin 1cm
\evensidemargin 0.5cm
\textwidth 15cm
\textheight 21cm
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}[theorem]{Definition}

\makeatletter
\def\thickhline{%
  \noalign{\ifnum0=`}\fi\hrule \@height \thickarrayrulewidth \futurelet
   \reserved@a\@xthickhline}
\def\@xthickhline{\ifx\reserved@a\thickhline
               \vskip\doublerulesep
               \vskip-\thickarrayrulewidth
             \fi
      \ifnum0=`{\fi}}
\makeatother

\newlength{\thickarrayrulewidth}
\setlength{\thickarrayrulewidth}{4\arrayrulewidth}


\def\W{$\mathbb{ W}$}
\def\R{\mathbb{ R}}
\def\S{\mathbb{ S}}
\def\I{\mathbb{ I}}
\makeindex


\title{Mbona Private Nature Reserve \\
        Small Scale Trout Hatcheries \\
        \vspace{2cm} 
        Mbona Fishing Committee \\
        \vspace{2cm}
        \today}

\author{
\htmladdnormallink           % Puts a hyperlink on to the author's name
{Hugh Murrell}{http://www.cs.ukzn.ac.za/~hughm}, Editor \\
\htmladdnormallink
{Pierre Olivier}{http://www.pierrecraft.co.za}, Technical \\
\htmladdnormallink
{Dennis Dyer}{http://www.mbona.co.za}, Chairman \\  
\vspace{1cm} \\
{\small\em \copyright none at all }}

 \date{ }
     \usepackage{draftwatermark}                   %%%   These four lines
      \SetWatermarkText{}   %%%   put a watermark
       \SetWatermarkScale{3}                          %%%  on all pages. Just omit them if 
        \SetWatermarkColor[rgb]{0.9,0,0}               %%%  it is not needed.
\begin{document}
\makeatletter\@openrightfalse
\maketitle
 \addcontentsline{toc}{chapter}{Contents}
\pagenumbering{roman}
\tableofcontents
\listoffigures
\listoftables


\chapter*{Preface}\normalsize
  \addcontentsline{toc}{chapter}{Preface}
\pagestyle{plain}

This manual is a basic guide to the running of a successful practice of 
small-scale trout farming. It summarises all the technical information that is 
important to know for small-scale trout production.

This document was created using 
{\htmladdnormallink {Leslie Lamport's}{ https://en.wikipedia.org/wiki/Leslie_Lamport}} document preparation system \LaTeX \cite{lamport}.
The content for the first chapter of this manual was mostly copied from a document  called {\it Small-scale rainbow trout farming} issued by 
the Food and Agriculture Organisation of the United Nations \cite{fao} which has been placed in
the public domain at 
{\htmladdnormallink{http://www.fao.org/docrep/015/i2125e/i2125e.pdf}{http://www.fao.org/docrep/015/i2125e/i2125e.pdf}} This {\bf fao} document has a copyright notice that encourages reproduction and dissemination of the material it contains with a statement that non-commercial uses will be authorized free of charge on request. 
So we must make a request to {\htmladdnormallink {copyright@fao.org}{ http://copyright@fao.org}} if we want to retain the material in the first chapter.

In later chapters, this FAO document is augmented using the local knowledge of:

\begin{itemize}
  \item[] {\bf Gareth Powell},  Mbona Conservation Officer, 2000-2018
  \item[] {\bf Dave Forsyth}, Mbona Manager, 2016-Present
  \item[] {\bf Gavin Chin}, retired Trout Hatchery manager.
\end{itemize}
  
Editing and oversight of this document was carried out by:
\begin{itemize}
\item[] {\bf Dr. Dennis Dyer}, Mbona fishing committee chairman, 2000-Present
\item[]  {\bf Bernard McDonald}, Local fisherman, historian and linguist.
\end{itemize}

The purpose of this manual is to guide the reader through the necessary technical information, 
and practical solutions to the day-to-day operation of a small-scale Rainbow Trout farm with
particular emphasis on the infrastructure setup and operating procedures at the Mbona Hatcheries.


\pagestyle{headings}
\pagenumbering{arabic}

\setcounter{chapter}{-1}
\include{ch0}
\include{ch1}
\include{ch2}
\include{ch3}
\include{ch4}
\include{ch5}
\include{ch6}
\include{chApp}
%\include{ch5}

\begin{thebibliography}{99}
  \addcontentsline{toc}{chapter}{Bibliography}
\bibitem{lamport} L. Lamport, {\bf \LaTeX \ A Document Preparation System}
Addison-Wesley, California 1986.
\bibitem{fao} András Woynarovich, György Hoitsy and Thomas Moth-Poulsen.
{\bf Small-scale rainbow trout farming}, FAO technical paper, 561, Rome, 2011,
available from {\htmladdnormallink{http://www.fao.org/docrep/015/i2125e/i2125e.pdf}{http://www.fao.org/docrep/015/i2125e/i2125e.pdf}}
\bibitem{waterpak}Bruce Finney, {\bf WaterPAK, a guide for irrigation management},
Australian Cotton Research and Development Corporation, 2013.
\bibitem{tic}{\bf Trout in the Classroom}, {\htmladdnormallink{http://www.troutintheclassroom.org/home}
{http://www.troutintheclassroom.org/home}}.

\end{thebibliography}

\include{index}
 \addcontentsline{toc}{chapter}{Index}
\end{document}
